{\Large \bf Analysing time course microarray data using Bioconductor: a case study using yeast2 Affymetrix arrays}

### 2012-07-12 paper.Rnw
Renamed bibtex file to references.bib. Stopped using the bmc_article.bst style file for references.
Removed the purl function call from the document. A couple of minor stylistic changes to the R code.

<<echo=FALSE>>=
opts_chunk$set(prompt = FALSE,fig.path='graphics/', fig.align='center',fig.lp="")
#knit_hooks$set()
options(width=90)
knit_hooks$set(
    par.nice = function(before, options, envir) {
        if (before && !is.null(options$par.mfrow)){
            par(mar=c(3,3,2,1),  mgp=c(2,0.4,0), tck=-.01,
                cex.lab=.95,cex.axis=0.9, las=1, mfrow=options$par.mfrow)

@@

{\small\begin{verbatim}
    AffyBatch object
    size of arrays=496x496 features (3163 kb)
    cdf=Yeast_2 (10928 affyids)
    number of samples=30
    number of genes=10928
    annotation=yeast2
\end{verbatim}}
\noindent If the Affymetrix microarray data sets have been downloaded into a single directory, then the \cc{.cel} files can be loaded into R using the \cc{ReadAffy} command.

@@

<<eval=TRUE,message=FALSE,warning=FALSE>>=
#Read in the mask file
s_cer = read.table('s_cerevisiae.msk', skip=2, 
stringsAsFactors=FALSE)
probe_filter = s_cer[[1]]
source('ExtractIDs.R')
c_df = ExtractIDs(probe_filter)
@
\noindent We also need to restrict the view of \cc{yeast.raw} to the $x-$ and $y-$coordinates of the \textit{S. cerevisiae} probesets in the \cc{cdf} environment by using
<<eval=TRUE,message=FALSE>>=
#Get the raw dataset for S. cerevisiae only
library(affy)
library(yeast2probe)
source('RemoveProbes.R')
cleancdf = cleancdfname(yeast.raw@cdfName)
RemoveProbes(probe_filter, cleancdf, 'yeast2probe') 
@
Note that the commands in \cc{RemoveProbes.R} are listed in Appendix A. Thus the attributes of \cc{yeast.raw}, obtained via
\cc{print(yeast.raw)}, are now
{\small\begin{verbatim}
    AffyBatch object
    size of arrays=496x496 features (3167 kb)
    cdf=Yeast_2 (5900 affyids)
    number of samples=30
    number of genes=5900
    annotation=yeast2
\end{verbatim}}
\noindent and the number of genes (actually probesets here) is 5,900 now that the \textit{S. pombe} probesets have been removed.

@@

Before any formal statistical analysis, it is important to check for
data quality. Initially, we might examine the perfect and mismatch
probe-level data to detect anomalies. Images of the first five arrays can be obtained using
<<s_fig1,eval=FALSE,tidy=FALSE>>=
for(i in 1:5){
    plot_title = paste('Strain: ', exp_fac$strain[i], 
        'Time: ', exp_fac$tps[i])
        d = exp_fac$data_order[i]
    image(yeast.raw[,d], main=plot_title) 
}
@

@@

<<F1,eval=TRUE, tidy=FALSE,dev="pdf",fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="A plot of the first two principal components. The red symbols correspond to the wild-type strain.", par.nice=TRUE>>=
#Plot of the first two principal components
plot(yeast.scores[,1], yeast.scores[,2], 
    xlab='PC 1', ylab='PC 2', 
    pch=rep(seq(1,5), 6), col=as.numeric(exp_fac$strain))
legend("bottomleft", pch=1:5, cex=0.6, 
    c('t 0', 't 60', 't 120', 't 180', 't 240'))
@

@@

% \begin{table}[h]
% \centering
% \begin{tabular}{@{}llll@{}}
% \toprule
% Package & Version & Package & Version \\
% \midrule
% affy & \Sexpr{packageDescription("affy")$Version}   & IRanges & \Sexpr{packageDescription("IRanges")$Version} \\
% affyPLM & \Sexpr{packageDescription("affyPLM")$Version}  & limma & \Sexpr{packageDescription("limma")$Version} \\
% AnnotationDbi & \Sexpr{packageDescription("AnnotationDbi")$Version} & Mfuzz & \Sexpr{packageDescription("Mfuzz")$Version} \\
% ArrayExpress & \Sexpr{packageDescription("ArrayExpress")$Version} & timecourse & \Sexpr{packageDescription("timecourse")$Version}  \\
% Biobase & \Sexpr{packageDescription("Biobase")$Version}  & yeast2probe & \Sexpr{packageDescription("yeast2probe")$Version} \\
% Biostrings & \Sexpr{packageDescription("Biostrings")$Version} & yeast2cdf & \Sexpr{packageDescription("yeast2cdf")$Version} \\
% gcrma & \Sexpr{packageDescription("gcrma")$Version}  & \\
% \bottomrule
% \end{tabular}
% \caption{List of packages from Bioconductor used in this paper.}\label{AT.1}
% \end{table}
% 
% <<echo=FALSE>>=
% #library()
% @
% 
% \begin{table}[h]
% \centering
% \begin{tabular}{@{}llll@{}}
% \toprule
% Package & Version & Package & Version \\
% \midrule
% bitops & \Sexpr{packageDescription("bitops")$Version}  & GeneNet & \Sexpr{packageDescription("GeneNet")$Version} \\
% caTools & \Sexpr{packageDescription("caTools")$Version} & gplots & \Sexpr{packageDescription("gplots")$Version} \\ 
% corpcor & \Sexpr{packageDescription("corpcor")$Version} & grid & \Sexpr{packageDescription("grid")$Version} \\
% e1071 & \Sexpr{packageDescription("e1071")$Version}  & gtools & \Sexpr{packageDescription("gtools")$Version}  \\
% fdrtool & \Sexpr{packageDescription("fdrtool")$Version} & MASS & \Sexpr{packageDescription("MASS")$Version} \\
% gdata & \Sexpr{packageDescription("gdata")$Version}  & longitudinal &
% \Sexpr{packageDescription("longitudinal")$Version}  \\
% \bottomrule
% 
% \end{tabular}
% \caption{List of packages from the CRAN repository used in this paper.}\label{AT.2}
% 
% \end{table}

@@

%{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
\bibliographystyle{bmc_article}  % Style BST file
\bibliography{refs}      % Bibliography file (usually '*.bib' )

@@

<<echo=FALSE>>=
library(knitr)
purl("paper.Rnw")
@

>>>>

<<echo=FALSE>>=
opts_chunk$set(prompt = FALSE,fig.path='graphics/', fig.align='center',fig.lp="")
#knit_hooks$set()
options(width=75)
knit_hooks$set(
    par.nice = function(before, options, envir) {
        if (before && !is.null(options$par.mfrow)){
            par(mar=c(3,3,2,1),  mgp=c(2,0.4,0), tck=-.01,
                cex.lab=.95,cex.axis=0.9, las=1, mfrow=options$par.mfrow)

@@

{\small\begin{verbatim}
    AffyBatch object
    size of arrays = 496x496 features (3163 kb)
    cdf = Yeast_2 (10928 affyids)
    number of samples = 30
    number of genes = 10928
    annotation = yeast2
\end{verbatim}}
\noindent If the Affymetrix microarray data sets have been downloaded into a single directory, then the \cc{.cel} files can be loaded into R using the \cc{ReadAffy} command.

@@

<<eval=TRUE,message=FALSE,warning=FALSE>>=
#Read in the mask file
s_cer = read.table('s_cerevisiae.msk', skip=2, 
stringsAsFactors=FALSE)
source('ExtractIDs.R')
c_df = ExtractIDs(s_cer[ ,1])
@
\noindent We also need to restrict the view of \cc{yeast.raw} to the $x-$ and $y-$coordinates of the \textit{S. cerevisiae} probesets in the \cc{cdf} environment by using
<<eval=TRUE,message=FALSE>>=
#Get the raw dataset for S. cerevisiae only
library(affy)
library(yeast2probe)
source('RemoveProbes.R')
cleancdf = cleancdfname(yeast.raw@cdfName)
RemoveProbes(s_cer[ ,1], cleancdf, 'yeast2probe') 
@
Note that the commands in \cc{RemoveProbes.R} are listed in Appendix A. Thus the attributes of \cc{yeast.raw}, obtained via
\cc{print(yeast.raw)}, are now
{\small\begin{verbatim}
    AffyBatch object
    size of arrays = 496x496 features (3167 kb)
    cdf = Yeast_2 (5900 affyids)
    number of samples = 30
    number of genes = 5900
    annotation = yeast2
\end{verbatim}}
\noindent and the number of genes (actually probesets here) is 5,900 now that the \textit{S. pombe} probesets have been removed.

@@

Before any formal statistical analysis, it is important to check for
data quality. Initially, we might examine the perfect and mismatch
probe-level data to detect anomalies. Images of the first five arrays can be obtained using
<<s_fig1,eval=FALSE,tidy=TRUE>>=
for(i in 1:5){
    plot_title = paste('Strain: ', exp_fac$strain[i], 
        'Time: ', exp_fac$tps[i])
        d = exp_fac$data_order[i]
    image(yeast.raw[,d], main=plot_title) 
}
@

@@

<<F1,eval=TRUE, tidy=FALSE,dev="pdf",fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="A plot of the first two principal components. The red symbols correspond to the wild-type strain.", par.nice=TRUE>>=
#Plot of the first two principal components
plot(yeast.scores[,1], yeast.scores[,2], xlab='PC 1', ylab='PC 2', 
    pch=rep(1:5, 6), col=as.numeric(exp_fac$strain))
legend("bottomleft", pch=1:5, cex=0.6, 
    c('t 0', 't 60', 't 120', 't 180', 't 240'))
@

@@

@@

%{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
\bibliographystyle{plain}  % Style BST file
\bibliography{references}      % Bibliography file (usually '*.bib' )

@@

% Not possible to call purl within a knitr document.
% <<echo=FALSE>>=
% library(knitr)
% purl("paper.Rnw")
% @
